---
title: "What Fictional Machine and Human Characters Do With Machine Vision Technologies"
subtitle: "Work in progress - please do not cite"
author: 
- Jill Walker Rettberg, University of Bergen, jill.walker.rettberg@uib.no
- Marianne Gunderson
- Linda Kronman
- Ragnhild Solberg
date: "`r Sys.Date()`" 
output:
  html_document:
    toc: true
    number_sections: true
    code_folding: hide
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
zotero: "Machine Vision in Everyday Life"
---

## To do:

1.  Go through all the scripts and rewrite now that I revised Characters
    to make it simpler.
2.  Add overview of other content analysis that's comparable
3.  Figure out how to get Zotero connected and add references
4.  Do more analysis of the character traits in combo with the verbs
5.  Figure out to what extend other team members will be involved - they
    already contributed some of the writing for the first version so
    probably need to be included, think about HOW.

# Introduction

This presentation explores the cultural imaginaries of machine vision
technologies as they are portrayed in contemporary science fiction,
digital art and videogames. How are the relationships between humans and
machines imagined in fictional situations and aesthetic contexts where
machine vision technologies occur? Our goal is to use a large-scale
analysis of fictional and artistic depictions of machine vision to to
understand how the distributed agency of human-technical assemblages is
imagined in our global cultural record.

We define machine vision as the registration, analysis and generation of
visual data by machines (Rettberg et al. 2019), and include technologies
such as facial recognition, optical implants, drones, surveillance
cameras and holograms in this. The project team has selected 500
creative works, primarily games, novels, movies, TV shows and artworks,
in which machine vision technologies are either described/shown or
actually used in the artwork. We have entered structured interpretations
of each work in the [Machine Vision in Art, Games and Narratives
Database](http://machine-vision.no/knowledgebase).

This is an analysis of the portrayal of human, machine and cyborg
characters in 500 narratives, games and artworks. While the first work
in the dataset, Arne Garborg's novel *Trætte Mænd*, is from 1981, 80% of
the works were published between 2011 and 2021, and just over half are
from 2016-2021. This paper focuses on the data about characters in the
works that are portrayed as directly interacting with machine vision
technologies.

This paper asks how fictional characters interactions with machine
vision technologies are portrayed.

# How to analyse agency

While there are many studies analysing gender representation in
videogames, movies and other media, most use content analysis to count
presence. We wanted to go beyond that and how different kinds of people
are imagined to be using machine vision in different ways.

Posthumanist and feminist new materialist theory argues that agency is
distributed between human and nonhuman actors. This is especially
pertinent with complex, algorithmically-driven technologies like machine
vision.

Our approach builds upon methodologies from literary studies and visual
analysis that rely on interpretation and close reading. The team read,
watched, played and experienced the 500 works, and entered their data
based on interpretative strategies learned through years of training.

Two alternative approaches are content analysis using a more traditional
quantitative coding or computational analysis of the textual material
itself.

Linguistic analysis of agency roles offers one approach to understanding
the agency of characters. Frazer Heritage (2021) used corpus linguistics
to analyse 350,000 words of transcribed text from ten videogames
published between 2012-2016, and found striking differences between the
actions described for male and female characters. While men were
connected to verbs like "discovered; faced; investigate; murdered;
showed", women were more likely to have collocates like "gave; made;
meet; took". The pronoun "he" was often collocated with violent words,
while there were no explicitly violent words in the list of words most
frequently collocated with "she". Heritage also found that while male
characters were "represented as enacting physical changes in the world,
female characters are the ones who have physical changes enacted upon
them.

Our approach was designed to work across multiple genres of creative
work, so relying on linguistic or visual analysis alone would not
suffice. We also wanted to preserve a key strength of literary analysis
and related methodologies: the hermeneutics of considering both details
and whole.

# Method

(Short general description of database, ref dataset and data in brief
paper).

We used R and the Tidyverse library to import data about characters and
machine vision situations. The characters data consists of a list of
characters that interact with machine vision technologies in the 500
works, and each character is described in the traits Species, Gender,
Race or Ethnicity, Age and Sexuality. We defined subsets consisting of
all the human characters and all the robots, where robots include all
kinds of software or hardware robots, cyborgs and AIs.

Situations are (DESCRIBE). We imported data about each situation
(situations.csv) and selected the data describing individual characters
and verbs that describe what each character *does* with machine vision
in the situation.

```{r, setup, message=FALSE, echo=FALSE}

library(tidyverse)
library(RColorBrewer)

#Import characters file (Characters.csv)
#define column types and factors
AllCharacters <- read_csv(
  "data/characters.csv",
  col_types = cols(
    CharacterID = col_integer(),
    Character = col_character(),
    Species = col_factor(levels = c(
      "Animal", "Cyborg", "Fictional", 
      "Human", "Machine", "Unknown")),
    Gender = col_factor(levels = c(
      "Female","Male","Non-binary or Other", "Trans Woman",
      "Unknown")),
    RaceOrEthnicity = col_factor(levels = c(
      "Asian", "Black", "White", "Person of Colour", "Immigrant", "Indigenous",
      "Complex", "Unknown")),
    Age = col_factor(levels = c(
      "Child", "Young Adult", "Adult", "Elderly", 
      "Unknown")),
    Sexuality = col_factor(levels = c(
      "Homosexual", "Heterosexual", "Bi-sexual", "Other",
      "Unknown")),
    IsGroup = col_logical(),
    IsCustomizable = col_logical()
  )
)

# Define Characters as the subset of AllCharacters that are not group characters or 
# customizable characters, and where the Species is Machine or Cyborg. 
# Convert "Unknown" values to NA. Create a new
# column called HumanOrMachine based on the Species column, and set all the machine 
# and cyborg characters to be "Robot", the humans to be "Human" and the other
# species to be "Other. Select relevant columns (leave out old Species) and 
# filter to only include humans and robots. 

Characters <- AllCharacters %>% 
        filter(IsGroup == FALSE & IsCustomizable == FALSE) %>% 
        na_if("Unknown") %>% 
        mutate(HumanOrMachine = recode(Species, 
                               "Machine" = "Robot",
                               "Cyborg" = "Robot",
                               "Human" = "Human",
                               .default = "Other")) %>% 
        select(Character, HumanOrMachine, Gender, Sexuality, 
               RaceOrEthnicity, Age) %>% 
        filter(HumanOrMachine == "Human" | HumanOrMachine == "Robot")


# To figure out what these characters actually do with the machine vision we need 
# to load data about the Situations in which they interact with machine vision
# technologies in the creative works in our sample.
# 
# The following code imports data about the Situations from situations.csv, 
# sets the column types, and also tells R to skip the columns we’re not going 
# to need for this analysis.

# NB characterID isn't in this export, fix later (add column back in)
# Also remove GenreID later

Situations <- read_csv(
  "data/situations.csv",
  col_types = cols(
    SituationID = col_integer(),
    Situation = col_skip(),
    Genre = col_character(),
    GenreID = col_skip(), 
    Character = col_character(),
    Entity = col_skip(),
    Technology = col_skip(),
    Verb = col_character()
  )
)

# Filter just the three main genres - since narratives have subgenres (Movie, 
# Novel, etc) there would be a lot of duplicate info if we kept them.

# The fifth row has an NA in the Character and CharacterID columns - that 
# means that there is no value there. The verb in the verb column belongs 
# to an Entity or a Technology, not to a Character. We need to delete all the 
# rows with missing data.

Situations <- drop_na(Situations, Character) # replace w/CharID later

# Now we combine the two dataframes using the CharacterID column as the 
# shared information.
Character_verbs <- merge(
        x = Situations, y = Characters, 
        by = "Character") %>% 
                        # replace w/CharID later
        select(Character, SituationID, Genre, Verb, HumanOrMachine, Gender, 
                RaceOrEthnicity, Age, Sexuality)

CreativeWorksFull <- read_csv("data/creativeworks.csv",
                              show_col_types = FALSE)
CreativeWorks <- CreativeWorksFull %>% 
        select(WorkTitle, Year, Genre, TechRef, Character) %>% 
        distinct()

Character_background <- merge(
        x = CreativeWorks, y = Characters,
        by = "Character")

Art <- Character_background %>% 
        filter(Genre == "Art")

Narrative <- Character_background %>% 
        filter(Genre == "Narrative")

Games <- Character_background %>% 
        filter(Genre == "Narrative")
                      

```

## Which technologies are used in which genres and by whom?

```{r, gender_tech}

# Selecting relevant columns and removing duplicates so the numbers are correct.
# This way each character occurs several times, but only once for each technology.
# Note that this looks at characters in relation to technologies referenced in the 
# whole work they are in, rather than in relation to the specific situation.

# I would like to order the bars according to the fill, and have been trying to do this 
# in "scripts/test-for-techref-char-graph.R"

Character_background %>% 
        filter(!is.na(TechRef)) %>% 
        select(TechRef, Character, Genre, Gender, Sexuality, Age, 
               RaceOrEthnicity, HumanOrMachine) %>% 
        distinct() %>%
        ggplot(aes(x = TechRef, fill = Gender)) +
        geom_bar(position = "fill") +
        scale_y_continuous(labels = scales::percent) +
        labs (x = "Technologies referenced in the work",
              y = "Gender of characters in works that reference these technologies",
              title ="Gender distribution of characters by technologies referenced in works") +
        theme(axis.text=element_text(size=8),
              title=element_text(size=10,face="bold")) +
        coord_flip() +
        facet_wrap(~Genre)
        

```

### Which technologies are most referenced?

```{r}

# See "scripts/tech_in_works.R" for attempts to make this for ALL tech, so with TechRef
# and TechUsed merged.

CreativeWorksFull %>% 
        select(WorkTitle, Year, Genre, Country, TechRef, TechUsed, Sentiment) %>% 
        distinct() %>%
        ggplot(aes(x = TechRef, fill = Genre)) +
        geom_bar()  +
        labs (x = "Technologies referenced in the work",
              y = "Count of works each tech is referenced in",
              title ="How often each technology is referenced in a work") +
        theme(axis.text=element_text(size=8),
              title=element_text(size=10,face="bold")) +
        coord_flip() +
        facet_wrap(~Genre)

# To do this with just Technology instead of TechRef vs TechUsed we could either
# - make one df x with just TechRef and another y with just TechUsed, rename 
# both columns Tech then vbind and 
# remove duplicates using distinct().
# - or use the technologies in situations instead and merge with creativeworks to 
# connect them to the WorkTitle.
        
```

### Characteristics of characters only in games and narratives

How many works actually have characters in the different genres?

```{r, art_characters, message=FALSE}
ArtWorksWithCharacters <- Character_background %>% 
        filter(Genre =="Art") %>% 
        select(WorkTitle, Character) %>% 
        distinct() %>% 
        group_by(WorkTitle) %>% 
        summarise(n = n()) %>% 
        count(WorkTitle)

GamesWithCharacters <- Character_background %>% 
        filter(Genre =="Game") %>% 
        select(WorkTitle, Character) %>% 
        distinct() %>% 
        group_by(WorkTitle) %>% 
        summarise(n = n()) %>% 
        count(WorkTitle)

NarrativesWithCharacters <- Character_background %>% 
        filter(Genre =="Narrative") %>% 
        select(WorkTitle, Character) %>% 
        distinct() %>% 
        group_by(WorkTitle) %>% 
        summarise(Total = n()) %>% 
        count(WorkTitle)

# To display the number of works in each genre with characters:
# cat("Number of artworks with characters: ", sum(ArtWorksWithCharacters$n))
# cat("Number of games with characters: ", sum(GamesWithCharacters$n))
# cat("Number of narratives with characters: ", sum(NarrativesWithCharacters$n))

WorksWithCharacters <- c(sum(ArtWorksWithCharacters$n), sum(GamesWithCharacters$n), 
                         sum(NarrativesWithCharacters$n))

CreativeWorks %>% 
        select(WorkTitle, Genre) %>% 
        distinct() %>% 
        group_by(Genre) %>% 
        summarise(TotalWorks = n()) %>% 
        add_column(WithChar = WorksWithCharacters) %>% 
        mutate(Percent = scales::percent(WithChar/TotalWorks))
        

```

OK, so which narratives and games DON'T have characters?

```{r, no_char_2nd_try}

# Find all the works that actually have characters.
CreativeWorks %>% 
        select(WorkTitle, Character, Genre) %>%  # only use WorkTitle and Character columns.
        distinct() %>%                    # remove duplicates
        mutate(HaveChar = ifelse(!is.na(Character), TRUE, FALSE)) %>% 
        select(WorkTitle, Genre, HaveChar) %>% 
                filter(WorkTitle != "Beholder" & WorkTitle != "I`m Alan Partrige" & WorkTitle != "The Matrix" 
               & WorkTitle !=  "The Expanse" & WorkTitle !=  "Six Wakes" 
               & WorkTitle !=  "Fall; or, Dodge in Hell: A Novel") %>%  # delete these when new export
        distinct() %>% 
        ggplot(aes(x = Genre, fill = HaveChar)) +
        geom_bar() + 
        geom_text(aes(
                label = ..count..), 
                stat = "count", 
                position = "stack", 
                size = 3,
                vjust = 1.2) +
        labs(x = "Genre of creative work",
             y = "Count of works in each genre",
              title ="Distribution of Characters by Genre") +
        scale_fill_manual(values=c("#999999", "#E69F00"), 
                       name="Work\nincludes\ncharacters",
                       labels=c("No", "Yes"))

# Delete line with all the titles after new export (errors in Jan 2022 export)

```

Most of the artworks don't have characters. Which games and narratives
don't have characters?

```{r}
CreativeWorks %>% 
        select(WorkTitle, Character, Genre) %>%  # only use WorkTitle and Character columns.
        distinct() %>%                    # remove duplicates
        mutate(HaveChar = ifelse(!is.na(Character), TRUE, FALSE)) %>% 
        filter(HaveChar == FALSE & Genre != "Art") %>% 
        filter(WorkTitle != "Beholder" & WorkTitle != "I`m Alan Partrige" & WorkTitle != "The Matrix" 
               & WorkTitle !=  "The Expanse" & WorkTitle !=  "Six Wakes" 
               & WorkTitle !=  "Fall; or, Dodge in Hell: A Novel") %>%  # delete these when new export
        select(WorkTitle, Genre) %>% 
        arrange(Genre)

# Errors in creativeworks export Jan 2022 - some extra rows with Character = NA. Have fixed 
# in database, need to get new export. Some records (like The Matrix) had extra rows with character 
# marked as NA. This caused them to show up as though they had no character although they actually did. 

```

Most artworks don't have characters in the sense that narratives and
games do. To capture agency in these, we used general "Entities", as we
called them, for general categories such as law enforcemnet, the creator
of the artwork or humans in general. These were not assigned specific
traits like gender and race. 8 out of 233 narratives have no characters,
and most of these have situations where only technologies and entities
are active. 13 out of 77 games have no characters, but all these have a
"User" who interacts with machine vision technologies. The User in a
game is the player when there is no clearly established player
character.

```{r}

Character_background %>% 
        filter(!is.na(TechRef)) %>% 
        select(TechRef, Character, Genre, Gender, Sexuality, Age, 
               RaceOrEthnicity, HumanOrMachine) %>% 
        filter(Genre == "Narrative" | Genre == "Game") %>% 
        distinct() %>%
        ggplot(aes(x = TechRef, fill = Gender)) +
        geom_bar() +
        scale_y_continuous(labels = scales::percent) +
        labs (x = "Technologies referenced in the work",
              y = "Gender of characters in works that reference these technologies") +
        coord_flip() +
        facet_wrap(~Genre)


   
```

## What kinds of characters interact with machine vision?

Here is a table to give us an overview. TO DO - Figure out how to
include the Döring & Poeschl 2019 data -- find a common key -- use
merge() - but I think that removes rows that don't have shared key, so
watch out for that? -- might want to split up humans vs robots first
then rbind all after adding new columns to simplify? - Are there other
stats I could include? - Consider installing a package to make tables
more beautiful and informative. E.g. these [tables with built-in bar
charts](https://glin.github.io/reactable/articles/cookbook/cookbook.html#bar-charts).

```{r, character_table}
# It would be better to write a function rather than all this repetition. I can't 
# 

HumanGender <- Characters %>% 
        filter(HumanOrMachine == "Human") %>% 
        count(Gender, sort = TRUE, name = "Count") %>% 
        mutate(Percent = round(Count/sum(Count)*100)) %>% 
        mutate(Variable = "Gender", .before = 1) %>%  
        mutate(Species = "Human", .before = 1)  %>% 
        rename(Value = Gender)

HumanRace <- Characters %>% 
        filter(HumanOrMachine == "Human") %>% 
        count(RaceOrEthnicity, sort = TRUE, name = "Count") %>% 
        mutate(Percent = round(Count/sum(Count)*100)) %>%  
        mutate(Variable = "Race or Ethnicity", .before = 1) %>% 
        mutate(Species = "Human", .before = 1) %>% 
        rename(Value = RaceOrEthnicity)

HumanAge <- Characters %>% 
        filter(HumanOrMachine == "Human") %>% 
        count(Age, sort = TRUE, name = "Count") %>% 
        mutate(Percent = round(Count/sum(Count)*100)) %>% 
        mutate(Variable = "Age", .before = 1) %>%  
        mutate(Species = "Human", .before = 1) %>% 
        rename(Value = Age)

HumanSexuality <- Characters %>% 
        filter(HumanOrMachine == "Human") %>% 
        count(Sexuality, sort = TRUE, name = "Count") %>% 
        mutate(Percent = round(Count/sum(Count)*100)) %>%  
        mutate(Variable = "Sexuality", .before = 1) %>% 
        mutate(Species = "Human", .before = 1) %>% 
        rename(Value = Sexuality)

RobotGender <- Characters %>% 
        filter(HumanOrMachine == "Robot") %>% 
        count(Gender, sort = TRUE, name = "Count") %>% 
        mutate(Percent = round(Count/sum(Count)*100)) %>%  
        mutate(Variable = "Gender", .before = 1) %>% 
        mutate(Species = "Robot", .before = 1) %>% 
        rename(Value = Gender)

RobotRace <- Characters %>% 
        filter(HumanOrMachine == "Robot") %>% 
        count(RaceOrEthnicity, sort = TRUE, name = "Count") %>% 
        mutate(Percent = round(Count/sum(Count)*100)) %>%  
        mutate(Variable = "Race or Ethnicity", .before = 1) %>% 
        mutate(Species = "Robot", .before = 1) %>% 
        rename(Value = RaceOrEthnicity)

RobotAge <- Characters %>% 
        filter(HumanOrMachine == "Robot") %>% 
        count(Age, sort = TRUE, name = "Count") %>% 
        mutate(Percent = round(Count/sum(Count)*100)) %>%  
        mutate(Variable = "Age", .before = 1) %>% 
        mutate(Species = "Robot", .before = 1) %>% 
        rename(Value = Age)

RobotSexuality <- Characters %>% 
        filter(HumanOrMachine == "Robot") %>% 
        count(Sexuality, sort = TRUE, name = "Count") %>% 
        mutate(Percent = round(Count/sum(Count)*100)) %>% 
        mutate(Variable = "Sexuality", .before = 1) %>% 
        mutate(Species = "Robot", .before = 1) %>% 
        rename(Value = Sexuality)

CharacterStats <- rbind(HumanGender, HumanRace, HumanAge, HumanSexuality, RobotGender, RobotRace, RobotAge, RobotSexuality)

# Display the CharacterStats table using kable in the knitr library. Probably better to use 
# a more fully featured table package, like Reactable (there are several others too). 
# 
# I set the NAs to show as Unknown in the table. Not sure whether this is the best option. 

library(knitr)
options(knitr.kable.NA = "Unknown")
kable(CharacterStats, caption = "An overview of the characteristics of the characters.")

```

This overview shows that for human characters,

Gender and age are explicitly described or shown in almost 90% of human
characters, and race or ethnicity are made explicit in 82% of the time.
Sexuality however, is only explicit in 39% of human characters. \[CHECK
DIFFERENCES GENRES\].

Robots do not have the cultural and biological ties that humans do, but
a vast majority (82%) of the robot characters in our dataset are still
explicitly assigned gender. Interestingly, more robots than humans are
assigned explicitly non-binary genders. Out of 540 humans, there are
only 3 non-binary and 2 transgender characters, so 1% of the characters.
There are only 101 robots in the dataset, but 5 are non-binary and one
transgender, which is equivalent to 6%.

```{r}
 Characters %>% 
        group_by(HumanOrMachine) %>%   
        summarise(Total = n())
```

Let's take a closer look at the non-binary and transgender characters:

TO DO: - Merge creativeworks and characters to show chars, title of
creative work, country, year, creator, genre - Look at [SEER
robot](https://machine-vision.no/character/seer-robot) - is it correct
to view this as non-binary? - might want to go more specific to
narratives/games?

## Non-binary characters

There are very few explicitly non-binary or transgender characters in
the dataset. Here is a complete list:

```{r}
Character_verbs %>% 
        filter(Gender == "Non-binary or Other" | Gender == "Trans Woman") %>% 
        select(Character, Genre, Gender, Sexuality)
```

Of these, Niki, Zizi

Niki is an explicitly non-binary teenager in Ted Anderson and Jen
Hickman's 2018 cyberpunk comic book *Moth & Whisper*. The publicly
non-binary musician and performance artist Mood Killer plays a role as a
sexually androgynous "cam boy" in their music video by that name. Zizi
is a character created by combining deepfakes of drag artists in Jake
Elwas's 2020 artwork *The Zizi Show*. Other characters tagged as
non-binary are more difficult to place. The "Red Girls" in the videogame
Nier:Automata are for instance represented visually as two traditionally
feminine girls, but they are instantiations of AIs \[RAGNHILD -
CHECK??\] and can multiply, and have a deep, booming male voice. While
it may seem a stretch to call these "non-binary" in the sense that is
usually used for humans, they do represent a queering of binary gender
that we wanted to capture.

There are also very few transgender characters: only three. The
representation of Chelsea Manning in *Probably Chelsea* is one. The
robot Paladin from Annalee Newitz's novel *Autonomous* (2017) is the
character where gender transitioning is most clearly represented.
Paladin is a combat robot with an organic human brain that assists with
facial recognition. At first, Paladin's operator, Eliasz, thinks of
Paladin as male, and when sexual attraction begins to arise between
them, this becomes a problem for Eliasz, who nourishes some homophobia.
Paladin does some research to find the origin of the human brain, and is
thrilled when she is told that the brain came from a woman. She begins
to see herself as female, and when she shares this new identity with
Eliasz, their sexual and romantic relationship flourishes. The third
transgender character in the dataset is Max Lao in the game
*Technobabylon* (2015). Here the transition is mentioned, but not
emphasised in the game.

We may have missed some non-binary or transgender characters, but
overall the lack of representation affirms other findings. In an
analysis of transcripts of dialogue and other text in ten video games,
Frank Heritage found no uses of "they" that referred to non-binary
characters[@heritage2021]. In a survey of games in Adrienne Shaw's LGBTQ
Video Game Archive, Utsch et.al. found that whiel LGBTQ characters in
video games have increased since the 1980s, but that the majority are
gay men, and LGBTQ characters of all kinds are far more prevalent in
Adventure and RPGs than in Fighting, Simulation or Shooter
games[@utsch2017].

```{r, character_race}
Characters %>% 
        ggplot() +
        geom_bar(aes(y = RaceOrEthnicity, fill = Gender)) +
        facet_grid(. ~ HumanOrMachine)

```

There are considerably more male than female humans. Interestingly the
genders are more even among the robots. You can also see that there
aren't many NAs for gender - most of the characters were represented in
a way that clearly expressed a specific gender. The machine characters
were an exception.

We used separate tags for "Machine" and "Cyborg" species because we were
interested in the agency of humans in comparison to non-humans, and the
cyborg seemed a potentially interesting hybrid categori. We tagged
robots and androids, as well as AI systems, as the "Machine" species,
and defined "Cyborg" as "A [@heritage2021a]being with both organic and
mechanical/machinic body parts."

TRYING TO MAKE A TABLE
<https://stackoverflow.com/questions/29549731/finding-percentage-in-a-sub-group-using-group-by-and-summarise>

### Comparing with the representation in ...

This compares (how) to Nicola Döring and Sandra Poeschl's 2019 content
analysis of the portrayal of sexual relationships between humans and
robots in fiction and non-fiction[@döring2019a], which is one of the few
studies to explore characteristics of relationships between humans and
technologies. Heritage uses NLP and finds that men are more often agents
SUMMARISE THIS BETTER - also Marianne's creepypasta [@gunderson2020]

\*\*NB: Not much point in including sexuality as a property of humans or
robots since it's actually a property of hte *relationship*, not the
individual.\*\*

Also need to actaully COMPARE with my data. Maybe put in same table,
somehow?

```{r sexbots}

# A simple visualisation of the distribution of gender and 
# sexuality as found in this paper:
# 
# Döring, N., Poeschl, S. Love and Sex with Robots: A Content Analysis of 
# Media Representations. Int J of Soc Robotics 11, 665–677 (2019). 
# https://doi.org/10.1007/s12369-019-00517-y
# 
# I manually copied the data for sexbots from [Table 3]
# (https://link.springer.com/article/10.1007/s12369-019-00517-y/tables/3) in 
# the paper, using the values form the column % under Fictional media. It's
# important to realise that the numbers are PERCENTAGES.

sexbots <- structure(list(
        Variables = c("Female", "Male", "Child", "Adult", "Elderly", 
                      "Heterosexual", "Homosexual", "Other"), 
        Human = c(43, 62, 27, 57, 2, 78, 21, 1), 
        Robot = c(46, 55, 23, 55, NA, 78, 21, 1), 
        Type = c("Gender", "Gender", "Age", "Age", "Age", "Sexuality", 
                 "Sexuality", "Sexuality")),
        row.names = c(NA, -8L), 
        class = c("tbl_df", "tbl", "data.frame"))

ggplot(data = sexbots, group = Type) +
        geom_col(mapping=aes(x = Variables,
                             y = Human,
                             fill = Type)) +
        ggtitle("Characteristics of humans who had sexual relationships with robots") +
        ylab("Percentage of sample") +
        xlab("Traits") +
        facet_wrap(~ Type, scale="free_x")
        
ggplot(data = sexbots, group = Type) +
        geom_col(mapping=aes(x = Variables,
                             y = Robot,
                             fill = Type)) +
        ggtitle("Characteristics of robots who had sexual relationships with humans") +
        ylab("Percentage of sample") +
        xlab("Traits") +
        facet_wrap(~ Type, scale="free_x")
                
```

## Human Sexuality

Why are there so few queer characters? The number of characters
explicitly identified as queer in the database is very low. \[give
numbers, explain categories\].

There are a lot of characters who are never described as having any
sexual or romantic feelings. These are in the category of characters any
sexuality or being involved in romantic relationships. This is often the
case with robots, like BB8 or R2D2 in Star Wars, or with child-like
characters, like Vanellope and Ralph in Ralph Breaks the Internet (check
title). Zombies and \[more examples\] are usually not see marked as
"Unknown/Not applicable". Most of these are not represented as having n
in a sexual light either. Groups of people also fall into this category.

Sometimes we see a character that in many ways is presented as
"culturally queer", but who is never shown in any sexual or romantic
situation. Shuri in the Marvel universe is an example of this. There is
a lot of speculation online that Shuri is lesbian or bisexual, and but
Marvel avoids showing her

We logged characters as heterosexual or homosexual if they were
mentioned as attracted to or in a sexual or romantic relationship with a
character of the opposite or same gender. If both genders, we logged
them as bisexual. Sometimes we may not be aware of all presentations of
the character \[e.g. Dr Who, shuri, characters in franchises -- might
not have read all -- so perhaps they have a same-sex relationship in
season X and so really aren't exclusively heterosexual.

Obviously problematic to lock people into fixed idnetities like this.
Fictional representations though, not actual people. We're interested in
how they're repreented. But unavoidably using lenses that can lead to
stereotypical depiction -- unfortunate. \[is there scholarship on this
that we could use here?\]

We have not logged the most important characters in the stories overall,
but only characters who are directly involved with machine vision
technologies. So it is possible that gay and queer characters are more
visible in the works as a whole, but queer characters are less likely to
be presented as engaging with machine vision technology. Unfortunately
we don't have data to prove or disprove this.

# Robot Sexuality

You may be surprised to find machine characters - robots and AIs, mostly
- have sexuality at all. As you can see below, the vast majority of
machine and cyborg characters *don't* have any explicit sexuality.

80% of the machine and cyborg characters are explicitly gendered. Out of
76 machine characters, there are only 15 where the gender is unknown or
not applicable (20%). 6 are explicitly non-binary, the others are fairly
evenly split between female-identifying (26) and male-identifying (29).
The distribution is similar for cyborg characters, with 3 out of 14
having unknown or not applicable gender, as well as 6 with male and 5
with female identities.

In contrast, only 10% of machine characters and 35% of cyborgs are
portrayed as have a romantic or sexual interest in another character.
All these have an assigned gender.

gender is usually assigned (how are demographics in comparison to
western countries?) but sexuality is far less frequently made explicit.
When sexuality is explicit, heterosexuality is heavily overrepresented
(check).

Far fewer characters are explicitly shown as homosexual or bi-sexual
than would be expected based on demographics in Western countries. Only
10 characters are tagged as homosexual. All are entered by Jill...which
suggests a methodological issue where other people haven't entered
sexuality? Or perhaps there just aren't homosexual characters in games
and art? A few of the seven bi-sexual characters are entered by others.
Lower visibility of queer characters -- this is typical and I can cite
something about it. Also research team may be defaulting to straight,
and there are a lot of Unknown/not applicable.

Compare gendering of machine characters to gendering of voice assistants
etc. (Cite Lai-Tze Fan if her book Unseen Hands is out or something else
she has done)

Let's make a new table called "Robots" that just has the robots and
cyborgs and AIs - the characters that are tagged as "Machine" or
"Cyborg".

*OK; I haven't figure that out yet but look at this ;)*

```{r}
Characters %>%
        drop_na(Sexuality) %>%
        ggplot() +
        geom_bar(aes(y=Sexuality, fill = Gender)) +
        facet_wrap(.~HumanOrMachine)


```

Maybe a table would be a better way of representing this data, so we can
see which machine characters are portrayed as having a sexual
preference.

## Oops

-   the followign analysis is for a version of this that no longer works
    sigh

Emily at the top of that list is the narrator of the novel [*Emily
Eternal*](https://machine-vision.no/creative-work/emily-eternal) by M.G.
Wheaton (2019). She is an artificial intelligence designed by MIT
researchers to feel empathy with humans, and she is experienced as
physically present by humans who wear an interface chip that allows
Emily to interface directly with their brains. In the first pages of the
novel this is explained to the reader:

> The chip allows me to manipulate Regina's senses of sight, smell,
> touch, and hearing. Her eyes tell her brain there's a Caucasian woman
> in her early thirties with brown hair, blue-green eyes, and a kind
> face sitting opposite her. Her ears tell her my voice has a mid-range
> pitch, not too low, not too high, with a slight New Englander's
> accent. Her nose tells her I use mostly fragrance-free soap, a
> kiwi-infused shampoo, no perfume, but a baby powder--scented
> antiperspirant. When I touch her hand or even embrace her, I come off
> as warm, upright but not rigid, and a good hugger. In return, the chip
> gives me unlimited access to her brain, including thoughts, memories,
> learned behaviors, hopes and dreams, worst fears, and all things in
> between. Utilizing bioalgorithms, I'm able to create a comprehensive
> neural map of an individual's mind that can then be used in a
> therapeutic context to help patients with their issues, large or
> small. (Wheaton, M. G. 2019. *Emily Eternal* (pp. 5-6). Grand Central
> Publishing. Kindle Edition.)

This is surprisingly typical: machine characters tend to be described as
very human-like, and one of the ways they are anthropomorphised is by
being assigned explicit gender, sexuality and race. The above excerpt
makes it easy to tag Emily for our dataset: her species (an AI, thus a
"Machine"), race, gender and age are all explicitly described. Her
sexuality is very clear later on in the novel: she is in love with
Jason:

> So yeah, he's my crush. Not entirely sure when it started (yeah,
> right---Emily's Crushing On Jason Protocol went online precisely nine
> weeks, four days, three hours, and fifty minutes ago...he was wearing
> a green cardigan), but it took a few hours of systems analysis to make
> certain it had developed naturally as part of my evolving
> socialization rather than another mental mouse maze my colleagues had
> dropped me into. That my emotional responses to it are not only real
> but ever-changing, as well as its possible-to-even-likely transience,
> is why I've kept it cloaked from my team. (p. 23).

```{r}
Characters %>% 
        filter(Character == "Emily")
```

## Why is there only one gay robot?

```{r}
Characters %>% 
        group_by(Sexuality) %>% 
        filter(!is.na(Sexuality)) %>% 
        ggplot() +
        geom_bar(aes(x=Gender, fill=Sexuality)) +
        facet_grid(. ~ HumanOrMachine) +
        scale_size_area(max_size = 7) +
        labs(title = "Gender and Sexuality of Machine and Cyborg Characters",
             y = "Count",
             x = "Gender") # +
 #       theme(plot.title = element_text(size = 10)) # including this line makes 
 #                                                   # the whole plot disappear!
```

# What do characters *do* with machine vision?

There are some notable differences in the verbs associated with
different genders in our dataset.

## Active verbs

```{r}

# Using the Tidyverse library the following code takes Character_verbs and then 
# does the following: 
# 1. Adds a column called VerbCount that gives a count of each verb 
# 2. Arranges the dataframe in decreasing order by VerbCount 
# 3. Filters the dataframe to only include rows where the VerbCount is more 
# than 20 
# 4. Filters the dataframe to only include rows where Gender isn’t unknown. 
# 5. Filters the dataframe to only include rows where the string “ing” is 
# detected in the Verb variable. 
# 6. Makes a plot where the visible objects are based on the Verb and Gender 
# columns. 
# 7. Uses geom_count to plot this - this is a geom that is good for discrete or 
# categorical values (like text rather than numbers) that shows a bigger or 
# smaller point (circle) depending on how many co-occurences each x and y value 
# have. 
# 8. Set the max size of the points to 7. 
# 9. Adds labels to the diagram.

Character_verbs %>%
  add_count(Verb, name = "VerbCount") %>% 
  arrange(desc(VerbCount)) %>%
  filter(VerbCount>20) %>% 
  filter(Gender != "Unknown") %>% 
  filter(str_detect(Verb, "ing")) %>% 
  ggplot(aes(y = Verb, x = Gender)) +
  geom_count(aes(colour = Gender), alpha = 0.8) +
  facet_grid(. ~ HumanOrMachine) +
  scale_size_area(max_size = 7) +
  labs(title = "Most frequent active verbs by gender",
       y = "Verbs",
       x = "Gender",
       size = "") +
   theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank())
```

## Passive Verbs

```{r}
Character_verbs %>%
  add_count(Verb, name = "VerbCount") %>% 
  arrange(desc(VerbCount)) %>%
  filter(VerbCount>20) %>% 
  filter(Gender != "Unknown") %>% 
  filter(str_detect(Verb, "ed")) %>% 
  ggplot(aes(y = Verb, x = Gender)) +
  geom_count(aes(colour = Gender), alpha = 0.8) +
  facet_grid(. ~ RaceOrEthnicity) +
  scale_size_area(max_size = 7) +
  labs(title = "Most frequent passive verbs by gender",
       y = "Verbs",
       x = "Gender",
       size = "") +
   theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank())
```

(the following is from SLSA version) Notably, the male characters are
portrayed as having far more active relationships with machine vision
technologies than characters of other genders are. This indicates that
women are more often portrayed as the object being acted upon by machine
vision, while men are more often portrayed as active subjects making use
of machine vision to further their own agency. While both men and women
in the dataset are often portrayed to be watching, women are more
frequently being watched, while men are a lot more likely to be the ones
doing the watching rather than being watched. Men are also more likely
to be killed by machine vision technologies, and there are more male
characters overall in our dataset.

Overall, this confirms typical gender stereotypes. Apparently, the
machine gaze is often aligned with the male gaze we know from decades of
cinema (Mulvey 1975), where men watch and people who are not men are
watched. Women do use machine vision technologies to watch others almost
as often as they are watched, though, and it's worth noting that
communicating is a very common use for both men and women, although more
so for women. Killing and being killed almost only happen to male
characters in our dataset.

## Do Robots and Humans interact with machine vision differently?

```{r}
# Fix this to work with new Characters script
# 
# Character_verbs %>%
  # Count how many occurances there are of each Verb. Make a new table
  # with a list of Verbs in one column and a second column indicating
  # how many times each Verb is used. Sort this in descending order.
#  count(Verb, sort = TRUE) %>%  
  # Take the first 20 rows of the new table that count() generated. 
#  slice(1:20) %>%
  # Now join that table to the Character_verbs table. R will drop all 
  # the rows with verbs NOT on the top-20 list.
#  left_join(Character_verbs) %>% 
  # Add a column renaming (recoding) Cyborgs to Machine so as to merge
  # Machine and Cyborg characters. Leave humans as humans and call 
  # everything else "Other" so we can easily get rid of it and just focus
  # on Machines and Humans.
#  mutate(HumanOrMachine = recode(Species, 
#                               "Machine" = "Machine",
#                               "Cyborg" = "Machine",
#                               "Human" = "Human",
#                               .default = "Other")) %>% 
# Remove all the Other species. Then remove all rows where Gender
# is Unknown.
#  filter(HumanOrMachine != "Other") %>% 
 # filter(Gender != "Unknown") %>% 
# Use ggplot() to generate a barchart.
#  ggplot() +
# the geom is barchart. Put Verb on the y axis, count will be on x.
# The fill colour depends on the Gender. (Note: set a colour palette
# later)
#  geom_bar(mapping = aes(y = Verb, fill=Gender)) +
# Make a grid with a separate graph for each variable in HumanOrMachine.
#  facet_grid(cols = vars(HumanOrMachine)) +
# Add labels.
#  labs(title = "What characters do with machine vision",
#       y = "The character is...",
#       x = "Number of occurences",
#       size = "") 
```

# To do - more visualisations to make

1.  Make a new column with values active or passive and use facet_grid()
    to make two different graphs based on that.
2.  Try putting the bars for machines/humans on top of each other (in
    same graph)
3.  Try making verbs proportional rather than a simple count so it's
    easier to compare machine/human.
4.  Add some charts showing how many characters of each gender etc.
5.  Could select just a few verbs (killing/killed? Scanning/Scanned?)
    and do more detailed graphs of the characters that use those verbs.
6.  Connect to technologies, entities in situations?
7.  Change colour palette - find the colourblind-friendly palette.
8.  Obviously connect to close readings.
9.  Create node and edge tables and feed into Gephi.
10. Test other kinds of graph.
11. What if the grid takes gender as main factor rather than species?
12. Bring genre art/game/narrative back and run the verbs based on that.

# Further research

Linking this data to existing crowdsourced and commercial datasets, or
to research datasets, could yield new insights. Some examples of
possible research questions: - Add star ratings from IMDB (movies) or
Goodreads (novels) or other services to find out whether there is a
correlation between high ratings and the representation of certain kinds
of human-technology relationship. -

UPDATE THIS - THIS IS FROM SLSA PAPER

Most research on fictional and artistic depictions of technologies uses
literary analysis and close reading as its method. We hope that using
digital humanities methodologies will allow us to say something about
the larger picture, and that it will generate new research questions
that can be explored in more nuance and detail through close readings.
Quantitative analysis can suggest overall tendencies and patterns, but
to really understand how machine vision technologies are portrayed in
art, games and narratives we will need to follow up the distant reading
with close readings. Combining exploratory analysis as in this paper
with "the beauty of small data," as Gabriel Recchia puts it (2020, 389),
will be a major part of our future research, and we have begun to
publish some of this, such as Linda Kronman's analysis of digital art
and machine vision (Kronman 2020).

# CRediT author statement

NB, check this - did everyone participate enough in this part of the
project? How to credit for database work etc if they didn't do much of
the analysis? How much writing is required to be included as Writing -
Original Draft? Maybe Jenny also should be included?

**Jill Walker Rettberg**: Conceptualization, Methodology, Investigation,
Data curation, Formal Analysis, Writing -- Original Draft, Writing --
Review & Editing, Visualization, Supervision, Project Administration,
Funding acquisition **Ragnhild Solberg**: Methodology, Investigation,
Data curation, Writing -- Original Draft **Linda Kronman**: Methodology,
Investigation, Data curation, Writing -- Original Draft **Marianne
Gunderson**: Methodology, Investigation, Data curation, Writing --
Original Draft **Stein Magne Bjørklund**: Software, Data curation **Linn
Heidi Stokkedal**: Methodology, Investigation, Data curation, Project
Administration **Kurdin Jacob**: Investigation, Data curation, Project
Administration **Gabriele de Seta**: Investigation, Project
Administration **Annette Markham**: Methodology.

# Acknowledgements

This research is part of the project Machine Vision in Everyday Life:
Playful Interactions with Visual Technologies in Digital Art, Games,
Narratives and Social Media, which has received funding from the
European Research Council (ERC) under the European Union's Horizon 2020
research and innovation programme (Grant agreement no. 771800).

# References
