---
title: "What Fictional Machine and Human Characters Do With Machine Vision Technologies"
subtitle: 
author: "Jill Walker Rettberg, Marianne Gunderson, Linda Kronman, Ragnhild Solberg"
date: "`r Sys.Date()`" 
output:
  html_document:
    toc: true
    number_sections: true
    code_folding: hide
    df_print: paged
---

## To do:
1. Go through all the scripts and rewrite now that I revised Characters to make it simpler.
2. Add overview of other content analysis that's comparable
3. Figure out how to get Zotero connected and add references
4. Do more analysis of the character traits in combo with the verbs
5. Figure out to what extend other team members will be involved - they already contributed some of the writing for the first version so probably need to be included, think about HOW. 

# Introduction

This presentation explores the cultural imaginaries of machine vision technologies as they are portrayed in contemporary science fiction, digital art and videogames. How are the relationships between humans and machines imagined in fictional situations and aesthetic contexts where machine vision technologies occur? Our goal is to use a large-scale analysis of fictional and artistic depictions of machine vision to to understand how the distributed agency of human-technical assemblages is imagined in our global cultural record.

We define machine vision as the registration, analysis and generation of visual data by machines (Rettberg et al. 2019), and include technologies such as facial recognition, optical implants, drones, surveillance cameras and holograms in this. The project team has selected 500 creative works, primarily games, novels, movies, TV shows and artworks, in which machine vision technologies are either described/shown or actually used in the artwork. We have entered structured interpretations of each work in the [Machine Vision in Art, Games and Narratives Database](http://machine-vision.no/knowledgebase).

The dataset includes data describing 77 games, 190 artworks and 233 narratives (in total 500 Creative Works) where machine vision technologies play an important role. This includes Creative Works produced between 1891 and 2021, but with a heavy emphasis on recent works. Only 20% are from 2010 or earlier. 80% of the Works are from 2011-2021, and just over half from 2016-2021. The Creative Works are from 59 different countries, with 78,6% from North America and Europe, and 21,4% from other parts of the world.

This is an analysis of the portrayal of human, machine and cyborg characters in 500 narratives, games and artworks. While the first work in the dataset, Arne Garborg's novel *Trætte Mænd*, is from 1981, 80% of the works were published between 2011 and 2021, and just over half are from 2016-2021. This paper focuses on the data about characters in the works that are portrayed as directly interacting with machine vision technologies.

This paper asks how fictional characters interactions with machine vision technologies are portrayed.

# Prior research
While there are many studies analysing gender representation in videogames, movies and other media, most use content analysis to count presence. We wanted to go beyond that and how different kinds of people are imagined to be using machine vision in different ways.

One 

# Method

(Short general description of database, ref dataset and data in brief paper).

We used R and the Tidyverse library to import data about characters and machine vision situations. The characters data consists of a list of characters that interact with machine vision technologies in the 500 works, and each character is described in the traits Species, Gender, Race or Ethnicity, Age and Sexuality. We defined subsets consisting of all the human characters and all the robots, where robots include all kinds of software or hardware robots, cyborgs and AIs.

Situations are (DESCRIBE). We imported data about each situation (situations.csv) and selected the data describing individual characters and verbs that describe what each character *does* with machine vision in the situation.

Please see

```{r, echo = FALSE, message = FALSE}

library(tidyverse)
library(RColorBrewer)

#Import characters file (Characters.csv)
#define column types and factors
AllCharacters <- read_csv(
  "data/characters.csv",
  col_types = cols(
    CharacterID = col_integer(),
    Character = col_character(),
    Species = col_factor(levels = c(
      "Animal", "Cyborg", "Fictional", 
      "Human", "Machine", "Unknown")),
    Gender = col_factor(levels = c(
      "Female","Male","Non-binary or Other", "Trans Woman",
      "Unknown")),
    RaceOrEthnicity = col_factor(levels = c(
      "Asian", "Black", "White", "Person of Colour", "Immigrant", "Indigenous",
      "Complex", "Unknown")),
    Age = col_factor(levels = c(
      "Child", "Young Adult", "Adult", "Elderly", 
      "Unknown")),
    Sexuality = col_factor(levels = c(
      "Homosexual", "Heterosexual", "Bi-sexual", "Other",
      "Unknown")),
    IsGroup = col_logical(),
    IsCustomizable = col_logical()
  )
)

# Define Characters as the subset of AllCharacters that are not group characters or 
# customizable characters, and where the Species is Machine or Cyborg. 
# Convert "Unknown" values to NA. Create a new
# column called HumanOrMachine based on the Species column, and set all the machine 
# and cyborg characters to be "Robot", the humans to be "Human" and the other
# species to be "Other. Select relevant columns (leave out old Species) and 
# filter to only include humans and robots. 

Characters <- AllCharacters %>% 
        filter(IsGroup == FALSE & IsCustomizable == FALSE) %>% 
        na_if("Unknown") %>% 
        mutate(HumanOrMachine = recode(Species, 
                               "Machine" = "Robot",
                               "Cyborg" = "Robot",
                               "Human" = "Human",
                               .default = "Other")) %>% 
        select(Character, HumanOrMachine, Gender, Sexuality, RaceOrEthnicity, Age) %>% 
        filter(HumanOrMachine == "Human" | HumanOrMachine == "Robot")


# To figure out what these characters actually do with the machine vision we need 
# to load data about the Situations in which they interact with machine vision
# technologies in the creative works in our sample.
# 
# The following code imports data about the Situations from situations.csv, 
# sets the column types, and also tells R to skip the columns we’re not going 
# to need for this analysis.

# NB characterID isn't in this export, fix later (add column back in)
# Also remove GenreID later

Situations <- read_csv(
  "data/situations.csv",
  col_types = cols(
    SituationID = col_integer(),
    SituationTitle = col_skip(),
    Genre = col_character(),
    GenreID = col_skip(), 
    Character = col_character(),
    Entity = col_skip(),
    Technology = col_skip(),
    Verb = col_character()
  )
)

# Filter just the three main genres - since narratives have subgenres (Movie, 
# Novel, etc) there would be a lot of duplicate info if we kept them.

# The fifth row has an NA in the Character and CharacterID columns - that 
# means that there is no value there. The verb in the verb column belongs 
# to an Entity or a Technology, not to a Character. We need to delete all the 
# rows with missing data.

Situations <- drop_na(Situations, Character) # replace w/CharID later

# Now we combine the two dataframes using the CharacterID column as the 
# shared information.
Character_verbs <- merge(
        x = Situations, y = Characters, 
        by = "Character") %>% 
                        # replace w/CharID later
        select(Character, SituationID, Genre, Verb, HumanOrMachine, Gender, 
                RaceOrEthnicity, Age, Sexuality)

                      

```

## What kinds of characters interact with machine vision?


```{r}
Characters %>% 
        ggplot() +
        geom_bar(aes(y = RaceOrEthnicity, fill = Gender)) +
        facet_grid(. ~ HumanOrMachine)

```

There are considerably more male than female humans. Interestingly the genders are more even among the robots. You can also see that there aren't many NAs for gender - most of the characters were represented in a way that clearly expressed a specific gender. The machine characters were an exception.

We used separate tags for "Machine" and "Cyborg" species because we were interested in the agency of humans in comparison to non-humans, and the cyborg seemed a potentially interesting hybrid categori. We tagged robots and androids, as well as AI systems, as the "Machine" species, and defined "Cyborg" as "A being with both organic and mechanical/machinic body parts."

TRYING TO MAKE A TABLE
https://stackoverflow.com/questions/29549731/finding-percentage-in-a-sub-group-using-group-by-and-summarise

### Comparing with the representation in ...

This compares (how) to Nicola Döring and Sandra Poeschl's 2019 content analysis of the portrayal of sexual relationships between humans and robots in fiction and non-fiction, which is one of the few studies to explore characteristics of relationships between humans and technologies.

```{r}
# The data for sexbots is taken from Table 3 in 
# 
# Döring, N., Poeschl, S. Love and Sex with Robots: A Content Analysis of 
# Media Representations. Int J of Soc Robotics 11, 665–677 (2019). 
# https://doi.org/10.1007/s12369-019-00517-y
# 
# 

                
```


## Human Sexuality

Why are there so few queer characters? The number of characters explicitly identified as queer in the database is very low. \[give numbers, explain categories\].

There are a lot of characters who are never described as having any sexual or romantic feelings. These are in the category of characters any sexuality or being involved in romantic relationships. This is often the case with robots, like BB8 or R2D2 in Star Wars, or with child-like characters, like Vanellope and Ralph in Ralph Breaks the Internet (check title). Zombies and \[more examples\] are usually not see marked as "Unknown/Not applicable". Most of these are not represented as having n in a sexual light either. Groups of people also fall into this category.

Sometimes we see a character that in many ways is presented as "culturally queer", but who is never shown in any sexual or romantic situation. Shuri in the Marvel universe is an example of this. There is a lot of speculation online that Shuri is lesbian or bisexual, and but Marvel avoids showing her

We logged characters as heterosexual or homosexual if they were mentioned as attracted to or in a sexual or romantic relationship with a character of the opposite or same gender. If both genders, we logged them as bisexual. Sometimes we may not be aware of all presentations of the character \[e.g. Dr Who, shuri, characters in franchises -- might not have read all -- so perhaps they have a same-sex relationship in season X and so really aren't exclusively heterosexual.

Obviously problematic to lock people into fixed idnetities like this. Fictional representations though, not actual people. We're interested in how they're repreented. But unavoidably using lenses that can lead to stereotypical depiction -- unfortunate. \[is there scholarship on this that we could use here?\]

We have not logged the most important characters in the stories overall, but only characters who are directly involved with machine vision technologies. So it is possible that gay and queer characters are more visible in the works as a whole, but queer characters are less likely to be presented as engaging with machine vision technology. Unfortunately we don't have data to prove or disprove this.

# Robot Sexuality

You may be surprised to find machine characters - robots and AIs, mostly - have sexuality at all. As you can see below, the vast majority of machine and cyborg characters *don't* have any explicit sexuality.

80% of the machine and cyborg characters are explicitly gendered. Out of 76 machine characters, there are only 15 where the gender is unknown or not applicable (20%). 6 are explicitly non-binary, the others are fairly evenly split between female-identifying (26) and male-identifying (29). The distribution is similar for cyborg characters, with 3 out of 14 having unknown or not applicable gender, as well as 6 with male and 5 with female identities.

In contrast, only 10% of machine characters and 35% of cyborgs are portrayed as have a romantic or sexual interest in another character. All these have an assigned gender.

gender is usually assigned (how are demographics in comparison to western countries?) but sexuality is far less frequently made explicit. When sexuality is explicit, heterosexuality is heavily overrepresented (check).

Far fewer characters are explicitly shown as homosexual or bi-sexual than would be expected based on demographics in Western countries. Only 10 characters are tagged as homosexual. All are entered by Jill...which suggests a methodological issue where other people haven't entered sexuality? Or perhaps there just aren't homosexual characters in games and art? A few of the seven bi-sexual characters are entered by others. Lower visibility of queer characters -- this is typical and I can cite something about it. Also research team may be defaulting to straight, and there are a lot of Unknown/not applicable.

Compare gendering of machine characters to gendering of voice assistants etc. (Cite Lai-Tze Fan if her book Unseen Hands is out or something else she has done)

Let's make a new table called "Robots" that just has the robots and cyborgs and AIs - the characters that are tagged as "Machine" or "Cyborg".

*OK; I haven't figure that out yet but look at this ;)*

```{r}
Characters %>% 
        ggplot() +
        geom_bar(aes(y = RaceOrEthnicity, fill = Gender)) +
        facet_grid(. ~ HumanOrMachine)

```
Here are the characters that *don't* have an explicit sexuality - huh, did I make a mistake? Are there MORE humans without an explicit sexuality?
```{r}
Characters %>% 
        filter(is.na(Sexuality)) %>% 
        group_by(HumanOrMachine) %>% 
        ggplot() +
        geom_bar(aes(x=Sexuality, fill = HumanOrMachine))+
        facet_grid(space="free_y")

```

Maybe a table would be a better way of representing this data, so we can see which machine characters are portrayed as having a sexual preference.



```{r}
# OK, this is old code - I changed Characters. Not comple.
 Characters %>% 
        group_by(HumanOrMachine) %>%   
        summarise(Total = n())
# That counts how many Robots and Humans there are but I'd have to join it back
# into the dataframe to use with the other variables so I can see percentages by
# going n/sum(Total)
      #  group_by(HumanOrMachine, Sexuality) %>% 
      #  filter(!is.na(Sexuality)) %>% 
      #  summarise(n()) # But here I actually want something like n/TotalInThatGroup
      #                 # Maybe I should set variables? TotalRobots = ...??                     

```

## Oops

- the followign analysis is for a version of this that no longer works sigh

Emily at the top of that list is the narrator of the novel [*Emily Eternal*](https://machine-vision.no/creative-work/emily-eternal) by M.G. Wheaton (2019). She is an artificial intelligence designed by MIT researchers to feel empathy with humans, and she is experienced as physically present by humans who wear an interface chip that allows Emily to interface directly with their brains. In the first pages of the novel this is explained to the reader:

> The chip allows me to manipulate Regina's senses of sight, smell, touch, and hearing. Her eyes tell her brain there's a Caucasian woman in her early thirties with brown hair, blue-green eyes, and a kind face sitting opposite her. Her ears tell her my voice has a mid-range pitch, not too low, not too high, with a slight New Englander's accent. Her nose tells her I use mostly fragrance-free soap, a kiwi-infused shampoo, no perfume, but a baby powder--scented antiperspirant. When I touch her hand or even embrace her, I come off as warm, upright but not rigid, and a good hugger. In return, the chip gives me unlimited access to her brain, including thoughts, memories, learned behaviors, hopes and dreams, worst fears, and all things in between. Utilizing bioalgorithms, I'm able to create a comprehensive neural map of an individual's mind that can then be used in a therapeutic context to help patients with their issues, large or small. (Wheaton, M. G. 2019. *Emily Eternal* (pp. 5-6). Grand Central Publishing. Kindle Edition.)

This is surprisingly typical: machine characters tend to be described as very human-like, and one of the ways they are anthropomorphised is by being assigned explicit gender, sexuality and race. The above excerpt makes it easy to tag Emily for our dataset: her species (an AI, thus a "Machine"), race, gender and age are all explicitly described. Her sexuality is very clear later on in the novel: she is in love with Jason:

> So yeah, he's my crush. Not entirely sure when it started (yeah, right---Emily's Crushing On Jason Protocol went online precisely nine weeks, four days, three hours, and fifty minutes ago...he was wearing a green cardigan), but it took a few hours of systems analysis to make certain it had developed naturally as part of my evolving socialization rather than another mental mouse maze my colleagues had dropped me into. That my emotional responses to it are not only real but ever-changing, as well as its possible-to-even-likely transience, is why I've kept it cloaked from my team. (p. 23).

```{r}
Characters %>% 
        filter(Character == "Emily")
```

## Why is there only one gay robot?

```{r}
Characters %>% 
        group_by(Sexuality) %>% 
        filter(!is.na(Sexuality)) %>% 
        ggplot() +
        geom_bar(aes(x=Gender, fill=Sexuality)) +
        facet_grid(. ~ HumanOrMachine) +
        scale_size_area(max_size = 7) +
        labs(title = "Gender and Sexuality of Machine and Cyborg Characters",
             y = "Count",
             x = "Gender") # +
 #       theme(plot.title = element_text(size = 10)) # including this line makes 
 #                                                   # the whole plot disappear!
```

# What do characters *do* with machine vision?

There are some notable differences in the verbs associated with different genders in our dataset.

## Active verbs

```{r}

# Using the Tidyverse library the following code takes Character_verbs and then 
# does the following: 
# 1. Adds a column called VerbCount that gives a count of each verb 
# 2. Arranges the dataframe in decreasing order by VerbCount 
# 3. Filters the dataframe to only include rows where the VerbCount is more 
# than 20 
# 4. Filters the dataframe to only include rows where Gender isn’t unknown. 
# 5. Filters the dataframe to only include rows where the string “ing” is 
# detected in the Verb variable. 
# 6. Makes a plot where the visible objects are based on the Verb and Gender 
# columns. 
# 7. Uses geom_count to plot this - this is a geom that is good for discrete or 
# categorical values (like text rather than numbers) that shows a bigger or 
# smaller point (circle) depending on how many co-occurences each x and y value 
# have. 
# 8. Set the max size of the points to 7. 
# 9. Adds labels to the diagram.

Character_verbs %>%
  add_count(Verb, name = "VerbCount") %>% 
  arrange(desc(VerbCount)) %>%
  filter(VerbCount>20) %>% 
  filter(Gender != "Unknown") %>% 
  filter(str_detect(Verb, "ing")) %>% 
  ggplot(aes(y = Verb, x = Gender)) +
  geom_count(aes(colour = Gender), alpha = 0.8) +
  facet_grid(. ~ HumanOrMachine) +
  scale_size_area(max_size = 7) +
  labs(title = "Most frequent active verbs by gender",
       y = "Verbs",
       x = "Gender",
       size = "") +
   theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank())
```

## Passive Verbs

```{r}
Character_verbs %>%
  add_count(Verb, name = "VerbCount") %>% 
  arrange(desc(VerbCount)) %>%
  filter(VerbCount>20) %>% 
  filter(Gender != "Unknown") %>% 
  filter(str_detect(Verb, "ed")) %>% 
  ggplot(aes(y = Verb, x = Gender)) +
  geom_count(aes(colour = Gender), alpha = 0.8) +
  facet_grid(. ~ RaceOrEthnicity) +
  scale_size_area(max_size = 7) +
  labs(title = "Most frequent passive verbs by gender",
       y = "Verbs",
       x = "Gender",
       size = "") +
   theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank())
```


(the following is from SLSA version) 
Notably, the male characters are portrayed as having far more active relationships with machine vision technologies than characters of other genders are. This indicates that women are more often portrayed as the object being acted upon by machine vision, while men are more often portrayed as active subjects making use of machine vision to further their own agency. While both men and women in the dataset are often portrayed to be watching, women are more frequently being watched, while men are a lot more likely to be the ones doing the watching rather than being watched. Men are also more likely to be killed by machine vision technologies, and there are more male characters overall in our dataset.

Overall, this confirms typical gender stereotypes. Apparently, the machine gaze is often aligned with the male gaze we know from decades of cinema (Mulvey 1975), where men watch and people who are not men are watched. Women do use machine vision technologies to watch others almost as often as they are watched, though, and it's worth noting that communicating is a very common use for both men and women, although more so for women. Killing and being killed almost only happen to male characters in our dataset.

## Do Robots and Humans interact with machine vision differently?

```{r}
# Fix this to work with new Characters script
# 
# Character_verbs %>%
  # Count how many occurances there are of each Verb. Make a new table
  # with a list of Verbs in one column and a second column indicating
  # how many times each Verb is used. Sort this in descending order.
#  count(Verb, sort = TRUE) %>%  
  # Take the first 20 rows of the new table that count() generated. 
#  slice(1:20) %>%
  # Now join that table to the Character_verbs table. R will drop all 
  # the rows with verbs NOT on the top-20 list.
#  left_join(Character_verbs) %>% 
  # Add a column renaming (recoding) Cyborgs to Machine so as to merge
  # Machine and Cyborg characters. Leave humans as humans and call 
  # everything else "Other" so we can easily get rid of it and just focus
  # on Machines and Humans.
#  mutate(HumanOrMachine = recode(Species, 
#                               "Machine" = "Machine",
#                               "Cyborg" = "Machine",
#                               "Human" = "Human",
#                               .default = "Other")) %>% 
# Remove all the Other species. Then remove all rows where Gender
# is Unknown.
#  filter(HumanOrMachine != "Other") %>% 
 # filter(Gender != "Unknown") %>% 
# Use ggplot() to generate a barchart.
#  ggplot() +
# the geom is barchart. Put Verb on the y axis, count will be on x.
# The fill colour depends on the Gender. (Note: set a colour palette
# later)
#  geom_bar(mapping = aes(y = Verb, fill=Gender)) +
# Make a grid with a separate graph for each variable in HumanOrMachine.
#  facet_grid(cols = vars(HumanOrMachine)) +
# Add labels.
#  labs(title = "What characters do with machine vision",
#       y = "The character is...",
#       x = "Number of occurences",
#       size = "") 
```

# To do - more visualisations to make
1. Make a new column with values active or passive and use facet_grid() to make two different graphs based on that.
2. Try putting the bars for machines/humans on top of each other (in same graph)
3. Try making verbs proportional rather than a simple count so it’s easier to compare machine/human.
4. Add some charts showing how many characters of each gender etc.
5. Could select just a few verbs (killing/killed? Scanning/Scanned?) and do more detailed graphs of the characters that use those verbs.
6. Connect to technologies, entities in situations?
7. Change colour palette - find the colourblind-friendly palette.
8. Obviously connect to close readings.
9. Create node and edge tables and feed into Gephi.
10. Test other kinds of graph.
11. What if the grid takes gender as main factor rather than species?
12. Bring genre art/game/narrative back and run the verbs based on that.

# Further research

UPDATE THIS - THIS IS FROM SLSA PAPER

Most research on fictional and artistic depictions of technologies uses literary analysis and close reading as its method. We hope that using digital humanities methodologies will allow us to say something about the larger picture, and that it will generate new research questions that can be explored in more nuance and detail through close readings. Quantitative analysis can suggest overall tendencies and patterns, but to really understand how machine vision technologies are portrayed in art, games and narratives we will need to follow up the distant reading with close readings. Combining exploratory analysis as in this paper with "the beauty of small data," as Gabriel Recchia puts it (2020, 389), will be a major part of our future research, and we have begun to publish some of this, such as Linda Kronman's analysis of digital art and machine vision (Kronman 2020).



# CRediT author statement

NB, check this - did everyone participate enough in this part of the project? How to credit for database work etc if they didn't do much of the analysis? How much writing is required to be included as Writing - Original Draft? Maybe Jenny also should be included?

**Jill Walker Rettberg**: Conceptualization, Methodology, Investigation, Data curation, Formal Analysis, Writing – Original Draft, Writing – Review & Editing, Visualization, Supervision, Project Administration, Funding acquisition **Ragnhild Solberg**: Methodology, Investigation, Data curation, Writing – Original Draft **Linda Kronman**: Methodology, Investigation, Data curation, Writing – Original Draft **Marianne Gunderson**: Methodology, Investigation, Data curation, Writing – Original Draft **Stein Magne Bjørklund**: Software, Data curation **Linn Heidi Stokkedal**: Methodology, Investigation, Data curation, Project Administration **Kurdin Jacob**: Investigation, Data curation, Project Administration  **Gabriele de Seta**: Investigation, Project Administration **Annette Markham**: Methodology.


# Acknowledgements

This research is part of the project Machine Vision in Everyday Life: Playful Interactions with Visual Technologies in Digital Art, Games, Narratives and Social Media, which has received funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme (Grant agreement no. 771800).

# References
