---
title: "Visualisations of which agents use specific groups of action"
subtitle: "Hacking and hacked"
author:
- Jill Walker Rettberg, University of Bergen, jill.walker.rettberg@uib.no
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    number_sections: yes
    code_folding: hide
    df_print: paged
  pdf_document:
    toc: yes

editor_options:
  markdown:
    wrap: 72
---

This notebook contains scripts that allow you to set a group of actions
(verbs in the dataset), to find synonyms of those actions that are also
in the database, and to create plots to show what kinds of agents use
those verbs.

# Setup

First we'll load the packages we're using.

```{r, setup, message=FALSE, echo=FALSE}
library(tidyverse)
library(syn)
```

## Set the actions 

In the next bit of code, we define the actions we're interested in.

```{r define_target_actions}

target_action <- c("Hacking", "Hacked", "Tricking", "Tricked", "Repurposing", "Repurposed", 
                   "Subverting", "Subverted", "Obfuscating", "Obfuscated", "Pranking", "Pranked")

# Define what you want the actions to be called in the plot.

name_of_target_actions <- "Hacking-related actions"
name_of_NON_target_actions <- "Not hacking-related"

plot_title <- "Which agents are hacking or hacked?"

```

Now import situations.csv.

```{r import_situations}
Situations <- read_csv("../data/situations.csv",
                       col_types = cols(
                               SituationID = col_integer(),
                               Situation = col_character(), 
                               Genre = col_character(),
                               Character = col_character(),
                               Entity = col_character(),
                               Technology = col_character(),
                               Verb = col_character()
                       )
)
```

## Look for synonyms

Let's see if there are any more actions we should be including as well
by using the syn() function to find synonyms to the verbs we set in
target_action. Note that the syn package is based on the Moby thesaurus
which is from 1996 and is often weak on verbs - it has none of the
"Hacking" verbs in it, for instance, but many "Flawed" verbs.

```{r find_synonyms}
syns(target_action)

```

When `syn()` returns `character(0)` that means that there are no
synonyms for the word in the thesaurus. If there are synonyms, you can
run the following code to redefine target_actions to include both the
verbs you entered at the start and the synonyms. The next lines convert
the verbs to title case and delete any that don't end in -ing or -ed and
finally those that aren't in our dataset.

```{r add_synonyms}

target_action <- c(target_action, unlist(lapply(target_action, syn)))
target_action <- str_to_title(
        target_action[str_detect(target_action, "ing")|str_detect(target_action, "ed")])
target_action <- target_action[target_action %in% Situations$Verb]
```

# Visualisations

Now we're ready to start examining the results.

## Which agents perform the target actions?

```{r plot_of_agents_and_target_actions, fig.height = 7}

Situations %>% 
        mutate(Target_action = case_when(Verb %in% target_action ~ name_of_target_actions,
                                  TRUE ~ name_of_NON_target_actions),
               Agent = case_when(!is.na(Entity) ~ "Entity",
                                 !is.na(Technology) ~ "Technology",
                                 !is.na(Character) ~ "Character")) %>% 
        filter(Target_action == name_of_target_actions) %>% 
        select(Genre, Verb, Entity, Technology, Agent) %>% 
        pivot_longer(!Verb, values_to = "value") %>%
        drop_na() %>% 
        ggplot(aes(x=fct_infreq(value), fill=Verb)) +
        geom_bar(aes(y = ..count..))+
        theme(axis.line = element_line(colour = "darkblue", 
                                       size = 1, 
                                       linetype = "solid")) +
        theme_minimal() +
        labs(title = plot_title,
             subtitle="",
             x = "") +
        coord_flip() +
        facet_wrap(~name, scales="free") 
```

## Same visualisation for agents that are NOT hacking or hacked

```{r plot_of_agents_NOT_doing_target_actions, fig.height = 7}

Situations %>% 
        mutate(Target_action = case_when(Verb %in% target_action ~ name_of_target_actions,
                                         TRUE ~ name_of_NON_target_actions),
               Agent = case_when(!is.na(Entity) ~ "Entity",
                                 !is.na(Technology) ~ "Technology",
                                 !is.na(Character) ~ "Character"),
               VerbTense = case_when(str_detect(Verb, "ing") ~ "Active",
                                     TRUE ~ "Passive")) %>% 
        filter(Target_action != name_of_target_actions) %>% 
        select(Genre, Verb, Entity, Technology, Agent, VerbTense) %>% 
        pivot_longer(!c(Verb, VerbTense), values_to = "value") %>%
        drop_na() %>% 
        ggplot(aes(x=fct_infreq(value), fill = VerbTense)) +
        geom_bar(aes(y = ..count..))+
        theme(axis.line = element_line(colour = "darkblue", 
                                       size = 1, 
                                       linetype = "solid")) +
        theme_minimal() +
        labs(title = "Distribution of agents in all the OTHER actions",
             subtitle="",
             x="") +
        coord_flip() +
                facet_wrap(~name, scales="free") 
```

## Compare the target actions to the rest of the actions in the dataset

The next plot compares the target actions - in this case actions
relating to hacking and tricking machine vision - to the rest of the
dataset. This is a "facet_grid" plot that shows a separate pane for each
genre by whether the verbs are active or passive. Each of these six
panes shows which agents, proportionally, do hacking-related and
non-hacking related actions.

```{r}

Situations %>% 
        mutate(Target_action = case_when(Verb %in% target_action ~ name_of_target_actions,
                                         TRUE ~ name_of_NON_target_actions),
               Agent = case_when(!is.na(Entity) ~ "Entity",
                                 !is.na(Technology) ~ "Technology",
                                 !is.na(Character) ~ "Character"),
               VerbTense = case_when(str_detect(Verb, "ing") ~ "Active",
                                     TRUE ~ "Passive")) %>% 
       #filter(Verb != "Drained") %>%  # remove # to manually remove any verbs that seem wrong
        select(Target_action, Agent, Genre, VerbTense) %>% 
        group_by(Agent, Target_action) %>% 
        ggplot(aes(x = Target_action, fill = Agent)) +
        geom_bar(position="fill") +
        labs(title = "Actions shown by genre and whether they are active (-ing) or passive (-ed)",
             subtitle="",
             x = "Actions",
             y = "Proportion of all interactions in dataset") +
        coord_flip() +
        facet_grid(VerbTense~Genre)

```

This is quite a dense visualisation to interpret!

The top left pane shows active verbs in artworks. So, in artworks,
technology is the agent in a little over half of all agent-action pairs
for the non-hacking related active verbs, but technology is hardly ever
an agent for the hacking-related actions. Entities are by far the most
common agent for hacking-related *active* verbs in artworks. However,
the bottom left pane shows that technology is by far the most common
agent for *passive* verbs in artworks. Since entities can (for instnace)
be users, governments or images, we'll need to dig deeper to understand
that better.

In narratives, characters are far more likely to be active agents with
hacking-related verbs, while technology is more likely to be passive.
The visualisation makes it very clear that hacking related actions are
quite different in this respect from the non-hacking related verbs.

## Same plot but by count instead of proportion

The last plot was proportional, so you don't notice that there are of
course WAY more non-hacking related actions than hacking-related
actions. This is the same plot, but showing absolute numbers instead of
the proportions.

```{r}

Situations %>% 
        mutate(Target_action = case_when(Verb %in% target_action ~ name_of_target_actions,
                                         TRUE ~ name_of_NON_target_actions),
               Agent = case_when(!is.na(Entity) ~ "Entity",
                                 !is.na(Technology) ~ "Technology",
                                 !is.na(Character) ~ "Character"),
               VerbTense = case_when(str_detect(Verb, "ing") ~ "Active",
                                     TRUE ~ "Passive")) %>% 
       #filter(Verb != "Drained") %>%  # remove # to manually remove any verbs that seem wrong
        select(Target_action, Agent, Genre, VerbTense) %>% 
        group_by(Agent, Target_action) %>% 
        ggplot(aes(x = Target_action, fill = Agent)) +
        geom_bar() +
        labs(title = "Actions shown by genre and whether they are active (-ing) or passive (-ed)",
             subtitle="",
             x = "Actions",
             y = "Proportion of all interactions in dataset") +
        coord_flip() +
        facet_grid(VerbTense~Genre)

```
