---
title: "What predicts whether an agent will take an active or passive role in a machine vision situation?"
subtitle: ""
author: "Jill Walker Rettberg"
output:
  html_document: 
    toc: yes
    number_sections: yes
    code_folding: hide
    fig_caption: yes
---

Trying to use machine learning to find clusters. Or, I could try to use machine learning to predict what kinds of character traits lead to cluster 1, 2 or 3.

Inspired by [this tutorial](https://www.r-bloggers.com/2022/02/beginners-guide-to-machine-learning-in-r-with-step-by-step-tutorial/) where they go through how to predict heart disease.

## Possible strategies:

1.  Make a matrix where the *target* variable (what we are trying to predict) is the cluster. I would need to do a thesaurus sort of some kind to sort ALL the verbs into one of the three clusters based on semantics to get the target. I might also have to make the values for character traits numeric? Perhaps just true or false

2.  Easier: predict the *target* of whether a verb is active or negative, since this is really easy to do.

3.  or use a different kind of machine learning not to predict but to detect patterns and find the clusters.

# Preparing the data

```{r setup}

suppressMessages(library(tidyverse))
suppressMessages(library(multiplex)) # think this is to make adjacency network?
                                        # actually no that's igraph


#Import characters file (../data/Characters.csv)
#define column types and factors
AllCharacters <- read_csv(
  "https://github.com/jilltxt/HumansRobotsAndMachineVision/raw/main/data/characters.csv",
  col_types = cols(
    CharacterID = col_integer(),
    Character = col_character(),
    Species = col_factor(levels = c(
      "Animal", "Cyborg", "Fictional", 
      "Human", "Machine", "Unknown")),
    Gender = col_factor(levels = c(
      "Female","Male","Non-binary or Other", "Trans Woman",
      "Unknown")),
    RaceOrEthnicity = col_factor(levels = c(
      "Asian", "Black", "Person of Colour", "White", "Immigrant", "Indigenous",
      "Complex", "Unknown")),
    Age = col_factor(levels = c(
      "Child", "Young Adult", "Adult", "Elderly", 
      "Unknown")),
    Sexuality = col_factor(levels = c(
      "Homosexual", "Heterosexual", "Bi-sexual", "Other",
      "Unknown")),
    IsGroup = col_logical(),
    IsCustomizable = col_logical()
  )
)

# Define Characters as the subset of AllCharacters that are not group characters or 
# customizable characters.
# 
# Convert "Unknown" values to NA. 
# 
# Select relevant columns.

Characters <- AllCharacters %>% 
        filter(IsCustomizable == FALSE) %>% 
        na_if("Unknown") %>% 
        select(Character, Species, Gender, Sexuality, 
               RaceOrEthnicity, Age)

# To figure out what these characters actually do with the machine vision we need 
# to load data about the Situations in which they interact with machine vision
# technologies in the creative works in our sample.
# 
# The following code imports data about the Situations from situations.csv, 
# sets the column types, and also tells R to skip the columns we’re not going 
# to need for this analysis.

# NB characterID isn't in this export, fix later (add column back in)
# Also remove GenreID later

Situations <- read_csv(
  "https://raw.githubusercontent.com/jilltxt/HumansRobotsAndMachineVision/main/data/situations.csv",
  col_types = cols(
    SituationID = col_integer(),
    SituationTitle = col_skip(),
    Genre = col_character(),
    Character = col_character(),
    Entity = col_character(),
    Technology = col_character(),
    Verb = col_character()
  )
)

# Add column stating whether verb is active or passive.
# Call this column "target" since this is the target or outcome we want to build
# a model to predict from the other variables. This variable is TRUE for active 
# verbs (ending in -ing) and FALSE for passive verbs (ending in -ed)
 
Situations <- Situations %>% 
        mutate(target = (str_detect(Verb, "ing"))) %>% 
        filter(!is.na(Verb))

# Filter just the three main genres - since narratives have subgenres (Movie, 
# Novel, etc) there would be a lot of duplicate info if we kept them.

# The fifth row has an NA in the Character and CharacterID columns - that 
# means that there is no value there. The verb in the verb column belongs 
# to an Entity or a Technology, not to a Character. We need to delete all the 
# rows with missing data.

Character_situations <- Situations %>% 
        select(SituationID, Genre, Character, Verb, target) %>% 
        filter(!is.na(Character))
        

# Now we combine the two dataframes using the CharacterID (for now: Character) 
# column as the shared information.
Character_verbs <- merge(
        x = Character_situations, y = Characters, 
        by = "Character") %>% 
                        # replace w/CharID later
        select(Character, SituationID, Genre, Verb, Species, Gender, 
                RaceOrEthnicity, Age, Sexuality, target)

head(Situations)

```

I want to predict whether an action will be passive or active based on whether it is a technology, entity or character, and on what kind of technology, entity of character it is. The final column in this table, "target", is TRUE if the verb is active (ends with -ing) and FALSE if it is passive (ends with -ed). We also have a subset of all the situation data which just has the characters with their traits. Here are the first few lines of this:

```{r show_character_verbs}

head(Character_verbs)
```

# Quick visual check of how the data is distributed

```{r distribution_entity_tech}
Situations %>% 
        select(Entity, Technology, target) %>% 
        gather() %>%
        drop_na() %>% 
  ggplot(aes(x=value), na.rm = TRUE) + 
  geom_bar(fill="steelblue", alpha=.7) +
  theme_minimal() +
        coord_flip() +
  facet_wrap(~key, scales="free")
```

Useful: There are many more active than passive actions. This could be to do with the way we have conceived of situations.

```{r fig.width=7,fig.height=7}

Situations %>% 
        mutate(IsCharacter=!is.na(Character)) %>% 
        mutate(ActiveVerb = target) %>% 
        select(Entity, IsCharacter, Technology, ActiveVerb, Genre) %>% 
        gather() %>%
        filter(!is.na(value)) %>% 
  ggplot(aes(x=value), na.rm = TRUE) + 
  geom_bar(fill="orange", alpha=.7) +
  coord_flip() +
  theme_minimal() +
  facet_wrap(~key, scale="free") +
        labs(title ="Number of actions across all situations for different kinds of agent")
```

TO DO: - make new df with year, country, include this data - one table with data on active/passive, entity/character/tech, genre - then another breaking down further.

Let's take a quick look at the characters.

```{r dataset_visual_check}

Characters %>% 
        select(Species, Gender, 
                RaceOrEthnicity, Age, Sexuality) %>% 
        gather() %>%
  ggplot(aes(x=value)) + 
  geom_bar(fill="orange", alpha=.7) +
  theme_minimal() +
        coord_flip() +
  facet_wrap(~key, scales="free")

```

## Another strategy: make the data quantitative

We can make a table listing a verb on each row, and the percentage of times it is used by a character who is female, male, human etc.

But what to do with this once we have it?

```{r charverb_percentage_table, fig.height=7}

Character_verbs_pct <- Character_verbs %>% 
        select(Verb, Gender, Species, RaceOrEthnicity, Age, Sexuality, Genre) %>% 
        pivot_longer(cols= -Verb,
                names_to = "variable", 
                values_to = "value") %>% 
        drop_na() %>% 
        group_by(Verb) %>%
        
        summarize(pct_female = mean(value == "Female", na.rm=TRUE),
                  pct.male = mean(value == "Male", na.rm=TRUE),
                  pct.trans = mean(value == "Trans Woman", na.rm=TRUE),
                  pct.human = mean(value == "Human", na.rm=TRUE),
                  pct.robot = mean(value == "Machine"|value=="Cyborg", na.rm=TRUE),
                  pct.animal = mean(value == "Animal", na.rm=TRUE),
                  pct.fictional = mean(value == "Fictional", na.rm=TRUE),
                  pct.asian = mean(value == "Asian", na.rm=TRUE),
                  pct.white = mean(value == "White", na.rm=TRUE),
                  pct.black = mean(value == "Black", na.rm=TRUE),
                  pct.PoC.other = mean(value == "Immigrant"|
                                               value == "Indigenous"|
                                               value == "Complex"|
                                               value == "Person of Colour", na.rm=TRUE),
                  pct.heterosexual = mean(value == "Heterosexual", na.rm=TRUE),
                  pct.homosexual = mean(value == "Homosexual", na.rm=TRUE),
                  pct.sexuality.other = mean(value == "Bi-sexual"|
                                                     value == "Other", na.rm=TRUE),
                  pct.child = mean(value == "Child", na.rm=TRUE),
                  pct.young.adult = mean(value == "Young Adult", na.rm=TRUE),
                  pct.adult = mean(value == "Adult", na.rm=TRUE),
                  pct.elderly = mean(value == "Elderly", na.rm=TRUE),
                  pct.narrative = mean(value == "Narrative", na.rm=TRUE),
                  pct.art = mean(value == "Art", na.rm=TRUE),
                  pct.game = mean(value == "Game", na.rm=TRUE))

# Now add the target (ActiveVerb is TRUE or FALSE)

Character_verbs_pct <- Character_verbs_pct %>%
        mutate(target = (str_detect(Verb, "ing"))) # add target column

# Tried making a correlation matrix, but it's clearly pointless, it just says stuff
# like "-0.76 correlation between male and female" which yeah, obviously they're
# negatively correlated and it's not -1 because there are other gender options. 
# Sigh. 
# 
# How do I even use this table? 

# cormat <- cor(Character_verbs_pct %>% keep(is.numeric))

# cormat %>% 
#        as.data.frame %>% 
#        mutate(var2=rownames(.)) %>%
#  pivot_longer(!var2, values_to = "value") %>%
#  ggplot(aes(x=name,y=var2,fill=abs(value),label=round(value,2))) +
#  geom_tile() + geom_label() + xlab("") + ylab("") +
#  ggtitle("Correlation matrix of our predictors") +
#  labs(fill="Correlation\n(absolute):") 


```

# Likely predictors of whether the verb is active or passive

## Character traits and active/passive actions

To quickly get an idea of which character traits might predict whether a verb is active ("watching") or passive ("watched") we can run barcharts showing the distribution of each trait

```{r}
Character_verbs %>% select(Genre, Species, Gender, 
                RaceOrEthnicity, Age, Sexuality, target) %>% 
        pivot_longer(!target, values_to = "value") %>%
        ggplot(aes(x=factor(value), fill=factor(target))) +
        scale_fill_manual(values=c("steelblue", "orangered1")) +
        geom_bar(position="fill", alpha=.7) +
        theme_minimal() +
        labs(fill="active verb:", title ="Association between character traits and active/passive verb",
             subtitle="(includes NA or missing data where the trait is unknown or not relevant") +
        coord_flip() +
        facet_wrap(~name, scales="free")
```

This already shows some likely predictors of how active the character is in their interaction with machine vision. Machines and cyborgs are a little more likely to be active than humans, and fictional species and animals are significantly less likely. Characters in artworks seem to have less active roles than those in games and narratives. Adults are more active than children and young adults. There aren't big differences in race or ethnicity, although white and Asian characters are a little more likely to haev active roles than Black characters, people of colour, indigenous characters and those with complex backgrounds. There are only three characters classified as immigrants, but their roles are all passive. Gender doesn't seem to have much effect, except trans women appear more active - but there are only three trans women (and no trans men) in the dataset.

## Character traits and active/passive actions - removing NAs (Unknowns)

NA means that data isn't available. Let's remove those:

```{r}
Character_verbs %>% select(Genre, Species, Gender, 
                RaceOrEthnicity, Age, Sexuality, target) %>% 
        pivot_longer(!target, values_to = "value") %>%
        drop_na() %>% 
        ggplot(aes(x=factor(value), fill=factor(target))) +
        scale_fill_manual(values=c("steelblue", "orangered1")) +
        geom_bar(position="fill", alpha=.7)+
        theme( axis.line = element_line(colour = "darkblue", 
                      size = 1, linetype = "solid")) +
        theme_minimal() +
        labs(fill="Verb is active:", title ="Association between character traits and active/passive verb",
             subtitle="(NAs removed)") +
        coord_flip() +
        facet_wrap(~name, scales="free")
```

## Entities, technologies and active/passive actions

```{r entities_tech_and_target}
Situations %>% 
        mutate(IsCharacter=!is.na(Character)) %>% 
        select(Entity, Technology, target) %>% 
        pivot_longer(!target, values_to = "value") %>%
        drop_na() %>% 
        ggplot(aes(x=factor(value), fill=factor(target))) +
        scale_fill_manual(values=c("steelblue", "orangered1")) +
        geom_bar(position="fill", alpha=.7)+
        theme( axis.line = element_line(colour = "darkblue", 
                      size = 1, linetype = "solid")) +
        theme_minimal() +
        labs(fill="Verb is active:", 
             xlab=NULL,
             ylab="Proportion of active to passive interactions",
             title ="Association between entities, technologies and active/passive verb") +
        coord_flip() +
        facet_wrap(~name, scales="free")
```

```{r random_70%_of_dataset}

set.seed(2022)
split <- sample(1:nrow(Character_verbs), as.integer(0.7*nrow(Character_verbs)), F)
train <- Character_verbs[split,]
test <- Character_verbs[-split,]

```

# **Partition data into training and test datasets**
