---
title: "Exploring the creative works"
subtitle: "Data visualisations based on creativeworks.csv from the Database of Machine Vision in Art, Games and Narratives"
author: "Jill Walker Rettberg"
output:
  html_document: 
    toc: yes
    number_sections: yes
    code_folding: hide
    fig_caption: yes
---

This is a collection of visualisations showing the general distribution of the creative works and so on.

Just started it March 2, 2022, more to come. 

```{r setup}

library(tidyverse)

creativeworks <- read_csv(
        "https://github.com/jilltxt/HumansRobotsAndMachineVision/raw/main/data/creativeworks.csv",
        col_types = cols(
                WorkID = col_integer(),
                WorkTitle = col_character(),
                Sentiment = col_factor(levels = c(
                        "Exciting", "Flawed", "Helpful", "Neutral", "Wondrous",
                        "Hostile","Oppressive", "Alien", "Creepy", "Subversive", 
                        "Dangerous",  "Intrusive", "Empowering", "Protective", 
                        "Intimate", "Misleading", "Fun", "Overwhelming", 
                        "Prosocial", "Disgusting")),
                Topic = col_factor(levels = c(
                        "Nudity", "Social Media", "Romantic relationship", "Climate Change", 
                          "Dystopian", "Horror", "Robots/androids", "Surveillance", "Automation", 
                          "City", "Labour", "War", "Identity", "AI", "Animals", "Consciousness", 
                          "Nature", "Companionship", "Competition", "Playful", "Family", 
                          "Free will", "Physical violence", "Crime", "Hacking", "Conflict", 
                          "Empathy", "Utopian", "Race", "Sex", "Cyborgs", "Inequality", 
                          "Economy", "Grief", "Autonomous vehicles", "Gender")),
                TechRef= col_factor(levels = c(
                        "Holograms", "Augmented reality", "Ocular implant", 
                        "Emotion recognition", "Surveillance cameras", "AI", 
                        "Virtual reality", "Motion tracking", "Body scans", 
                        "Drones", "MicroscopeOrTelescope", "Biometrics", 
                        "Image generation", "Facial recognition", 
                        "Object recognition",  "3D scans", "Machine learning", 
                        "Filtering", "Deepfake", "Camera",  "Cameraphone", 
                        "Interactive panoramas", "Non-Visible Spectrum", "UGV",
                        "Webcams", "Satellite images")),
                TechUsed= col_factor(levels = c(
                        "Holograms", "Augmented reality", "Ocular implant", 
                        "Emotion recognition", "Surveillance cameras", "AI", 
                        "Virtual reality", "Motion tracking", "Body scans", 
                        "Drones", "MicroscopeOrTelescope", "Biometrics", 
                        "Image generation", "Facial recognition", 
                        "Object recognition",  "3D scans", "Machine learning", 
                        "Filtering", "Deepfake", "Camera",  "Cameraphone", 
                        "Interactive panoramas", "Non-Visible Spectrum", "UGV",
                        "Webcams", "Satellite images"))))

```


# Distribution of works by contintent
```{r continents}

library(countrycode)

creativeworks$Continent <- countrycode(sourcevar = creativeworks$Country, origin = "country.name",destination = "region")

creativeworks %>% 
        select(WorkID, WorkTitle, Genre, Country, Continent) %>% 
        distinct() %>% 
        ggplot(aes(x=Genre, fill=Continent)) +
        geom_bar() +
        geom_text(stat = "count", 
                  aes(label =..count..), 
                  position = position_stack(vjust = 0.5)) +
        labs(fill="", 
             title ="Creative works by genre and continent",
             subtitle = "Note that some works are associated with several countries so the sum is more than the total works in the dataset",
             y = "", 
             x = "") +
        theme_minimal()


```

# Or the continents the works are from by genre

```{r}
library(countrycode)

# Create new column called Region giving the region a country belongs to according
# to  the World Bank Development Indicators. 
# See https://cran.r-project.org/web/packages/countrycode/countrycode.pdf for details

creativeworks$Region <- countrycode(sourcevar = creativeworks$Country, origin = "country.name",destination = "region")


creativeworks %>% 
        select(WorkID, WorkTitle, Genre, Country, Region) %>% 
        distinct() %>% 
        ggplot(aes(x=Region, fill=Genre)) +
        geom_bar() +
        geom_text(stat = "count", 
                  aes(label =..count..), 
                  position = position_stack(vjust = 0.5)) +
        labs(fill="", 
             title ="Creative works by region and genre",
             subtitle = "Regions as defined in the World Bank Development Indicators",
             y = "", 
             x = "") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle=45, hjust = 1, vjust = 1))
```

```{r}

#same but with UN region instead of world bank as above

creativeworks$UN_region <- countrycode(sourcevar = creativeworks$Country, origin = "country.name",destination = "un.region.name")

creativeworks %>% 
        select(WorkID, WorkTitle, Genre, Country, UN_region) %>% 
        distinct() %>% 
        ggplot(aes(x=UN_region, fill=Genre)) +
        geom_bar() +
        geom_text(stat = "count", 
                  aes(label =..count..), 
                  position = position_stack(vjust = 0.5)) +
        labs(fill="", 
             title ="Creative works by continent and genre (UN regions)",
             subtitle = "Continents defined by United National regions",
             y = "", 
             x = "") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle=45, hjust = 1, vjust = 1))
```
# Agiain with UN intermediate regions

```{r}
#same but with UN region instead of world bank as above

creativeworks$Region <- countrycode(sourcevar = creativeworks$Country, origin = "country.name",destination = "un.regionsub.name")

levels(creativeworks$Region) <- c("Northern Europe", "Eastern Europe",  "Western Europe", "Southern Europe", "Northern America", "Latin America and the Caribbean", 
"Northern Africa",   "Sub-Saharan Africa", 
"Eastern Asia","Central Asia","Western Asia","South-eastern Asia","Southern Asia", "Australia and New Zealand",   NA)

creativeworks %>% 
        select(WorkID, WorkTitle, Genre, Country, Region) %>% 
        distinct() %>% 
        ggplot(aes(x=factor(Region), fill=Genre)) +
        geom_bar() +
        geom_text(stat = "count", 
                  aes(label =..count..), 
                  position = position_stack(vjust = 0.5)) +
        labs(fill="", 
             title ="Creative works by continent and genre",
             subtitle = "United Nations sub-regions",
             y = "", 
             x = "") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle=45, hjust = 1, vjust = 1))
```



# Rather dodgy run-through

ignore this - doesn't remove duplicates so narratives are vastly exaggerated. Probably can't use facet wrap like this. 

```{r}
creativeworks %>% 
        select(Genre, Sentiment, Topic, Continent, TechRef, TechUsed) %>% 
        gather() %>%
        drop_na() %>% 
  ggplot(aes(x=value), na.rm = TRUE) + 
  geom_bar(fill="steelblue", alpha=.7) +
  theme_minimal() +
        coord_flip() +
  facet_wrap(~key, scales="free")
```
